Negative Perceptions of Outsourcing to Artificial Intelligence

### 2025-04-30 manuscript.qmd
Editing manuscript and preparing for submission

---
title: "Negative Perceptions of Outsourcing to Artificial Intelligence"
shorttitle: "Outsourcing to AI"
author:
  - name: Scott Claessens
    corresponding: false
    orcid: 0000-0002-3562-6981
    email: scott.claessens@gmail.com
    affiliations:
      - name: University of Kent
        department: School of Psychology
        address: Keynes College
        city: Canterbury
        country: UK
        postal-code: CT2 7NP
    role:
      - data curation
      - formal analysis
      - investigation
      - methodology
      - visualization
      - writing
      - editing
  - name: Pierce Veitch
    corresponding: false
    orcid: 0009-0005-3364-7470
    email: pv201@kent.ac.uk
    affiliations:
      - name: University of Kent
        department: School of Psychology
        address: Keynes College
        city: Canterbury
        country: UK
        postal-code: CT2 7NP
    role:
      - formal analysis
      - methodology
  - name: Jim Everett
    corresponding: true
    orcid: 0000-0003-2801-5426
    email: j.a.c.everett@kent.ac.uk
    affiliations:
      - name: University of Kent
        department: School of Psychology
        address: Keynes College
        city: Canterbury
        country: UK
        postal-code: CT2 7NP
    role:
      - conceptualization
      - funding acquisition
      - methodology
      - supervision
      - editing
author-note:
  status-changes: 
    affiliation-change: ~
    deceased: ~
  disclosures:
    study-registration: ~
    data-sharing: All data and original code can be found on GitHub at <https://github.com/ScottClaessens/outsourcing>.
    related-report: ~
    conflict-of-interest: The authors have no conflicts of interest to disclose.
    financial-support: This work was supported by ...
    gratitude: ~
    authorship-agreements: ~
abstract: >
  As artificial intelligence (AI) tools become increasingly integrated into 
  daily life, people are beginning to outsource not only professional tasks but
  also socio-relational ones. Large language models like ChatGPT can generate
  wedding vows, speeches, and personal messages, raising questions about how
  individuals who use AI for such tasks are perceived by others. In this paper,
  we conduct five pre-registered studies with British participants (*n* = 3649)
  to understand how people view those who outsource tasks to AI, and how this
  depends on how socio-relational the task is, whether AI is used as a tool or
  fully delegated to, and the acknowledgment of the AI use. We find negative
  perceptions of outsourcing, particularly for socio-relational tasks, and
  demonstrate that this does not only make us think more negatively about the
  outsourcer, but the outsourced work itself. We then turn to examine potential
  mechanisms underlying negative perceptions of outsourcing, investigating the
  roles of effort, perceived authenticity, and the reasons for using AI. Our
  research highlights the way that AI use shapes our perceptions of people,
  raising key philosophical questions about praise, blame, and authenticity in a
  world filled with AI-mediated interactions.
keywords: [artificial intelligence, character evaluation, outsourcing, trust]
bibliography: bibliography.bib
engine: knitr
format:
  apaquarto-pdf: default
execute:
  echo: false
  warning: false
  error: false
crossref:
  custom:
    - kind: float
      key: suppfig
      latex-env: suppfig
      reference-prefix: Supplementary Figure
      space-before-numbering: true
      latex-list-of-description: Supplementary Figure
      caption-location: bottom
    - kind: float
      key: supptbl
      latex-env: supptbl
      reference-prefix: Supplementary Table
      space-before-numbering: true
      latex-list-of-description: Supplementary Table
      caption-location: top
floatsintext: true
---

@@

In recent years, artificial intelligence (AI) technologies have emerged as 
useful tools to save time and effort across a wide variety of tasks. Large 
language models, such as ChatGPT, Copilot, and DeepSeek, have been shown to 
increase efficiency for professional tasks like writing [@Noy2023], project 
planning [@Barcaui2023], and programming [@Peng2023]. In a 2022 Pew survey of
the American public, time-saving and efficiency were listed among the top
reasons for excitement about the increased presence of AI in daily life
[@Rainie2022].

As outsourcing routine tasks to AI becomes more commonplace, people may begin to
use AI tools to efficiently complete more socio-relational tasks. Consider, for
example, writing wedding vows. While wedding vows are widely understood to be an
opportunity for couples to express their love in their own words, large language
models like ChatGPT will dutifully produce wedding vows upon request, without
necessarily requiring any information about the partner in question. In
characteristic ChatGPT style, the resulting text will likely be generic enough
to apply to anyone, yet well-written enough to read aloud at the altar. A recent
Reddit post from a disgruntled newlywed suggests that AI-outsourcing of this
kind is already happening [@miramar0].

Of course, there is nothing new about outsourcing socio-relational tasks.
Ghostwriters have existed for years: in the 19th century, Victor Hugo, the
writer of *Les Misérables*, used to write love letters for others to earn extra
money [ref]. Today, the internet is abound with professional paid services for
writing wedding vows and personal speeches. But it remains unclear how people
who outsource tasks like these are perceived by others, particularly when the
tasks are outsourced to impersonal AI tools like ChatGPT.

There are several reasons to expect that people will be perceived negatively for
outsourcing tasks to AI. First, the effort put into a task is often used as a
signal of one's moral character [@Celniker2023; @Tissot2025]. People who free
ride on the efforts of others are perceived as less moral [@Cubitt2011], and
people's outputs are perceived as more valuable the more effort was ostensibly
put into them [@Kruger2004]. Second, people are given less credit when their
work was generated jointly with or entirely by AI instead of by themselves
[@Earp2024]. Third, from a virtue ethics perspective, people who outsource moral
tasks to AI skip the opportunity to cultivate virtuous habits and develop their
moral character [@Vallor2015]. These disparate lines of reasoning predict that
people will be perceived more negatively if they outsource tasks to AI, though
this has yet to be tested.

If people are perceived more negatively for outsourcing tasks to AI, there is 
also the question of how these negative perceptions might vary across different
types of tasks and different kinds of AI use. Will outsourcing wedding vows to
AI be perceived just as negatively as outsourcing a cover letter or a few lines
of computer code? Given that effort is thought to be a particularly important
cue for cooperative partner choice in social situations [@Celniker2023], we 
predict that outsourcing socio-relational tasks to AI will be perceived more 
negatively than routine or professional tasks.

It is also possible for people to outsource to AI in different ways. One can 
complete tasks by using AI as a collaborative tool, sculpting its answer over 
successive prompts and editing the resulting output by hand. Alternatively, one
can simply copy the AI's output verbatim. To the extent that the latter approach
requires less effort, we predict that "fully" outsourcing to AI will be
perceived more negatively. People can also differ in whether they are honest
about their use of AI or, in the words of the disgruntled newlywed, deceptively
"pass [the work] off as [their] own" [@miramar0]. Since people tend to dislike
and distrust those who are deceptive [@Tyler2006], it follows that people will
perceive deceptive AI-outsourcing more negatively.

Here, we test these predictions across five pre-registered experiments with 
British participants. In our initial pilot studies, we found that people who 
outsource a range of tasks to AI are perceived more negatively than people who
complete the tasks by themselves (see Supplementary Materials). In Study 1, we
expand on this by manipulating the type of outsourcing to AI (as a collaborative
tool vs. fully outsourcing) and whether people are honest or deceptive in their
usage of AI. In Study 2, we measure perceptions of both outsourcers and the
outsourced work. In Studies 3-5, we test potential mechanisms underlying
negative perceptions of outsourcing, including perceived effort, perceived
authenticity, and the reasons for using AI.

@@

Ethical approval was granted for all studies in this paper by the University of
Kent Psychology Research Ethics Panel. Participants in all studies provided
informed consent and were debriefed after the study.

@@

We fitted Bayesian multivariate multilevel cumulative-link ordinal models to the
data using the *brms* R package [@Burkner2017]. We modelled each character 
evaluation -- competence, warmth, morality, laziness, and trustworthiness -- as 
a separate response variable and included fixed effects for conditions, varying 
intercepts for participants, and varying intercepts and slopes for tasks. We 
used regularising priors for all parameters to impose conservatism on parameter 
estimates (see Supplementary Materials for full model specifications). All 
models converged normally ($\hat{R}$ ≤ 1.01).

@@

Across all five character evaluations, we found that fully outsourcing to AI 
(i.e., copying the AI output verbatim) was perceived more negatively than using
AI as a collaborative tool (@fig-treatments-study1; 
@tbl-treatment-diffs-study1). By contrast, we found that deception about AI
usage had specific negative effects on perceptions of morality and
trustworthiness: people who did not acknowledge their use of AI were perceived
as less moral and less trustworthy. We did not find any interaction effects
between full outsourcing and deception.

```{r, fig.height=6, fig.width=6}
#| label: fig-treatments-study1
#| fig-cap: Character Evaluations in Study 1
#| fig-align: center
#| apa-note: Participants in the control condition and four AI outsourcing conditions evaluated people in the scenarios on (a) competence, (b) warmth, (c) morality, (d) laziness, and (e) trustworthiness. Coloured points represent participant responses to the questions, jittered for easier viewing. Black points are estimated marginal means from the fitted model, pooling over participants and tasks. Black points and line ranges represent posterior medians and 95% credible intervals, respectively.

@@

```{r}
#| label: tbl-treatment-diffs-study1
#| tbl-cap: Pairwise Contrasts in Study 1 \vspace{20pt}
#| apa-note: Numbers reflect differences in marginal means on a 7-point Likert scale, pooling over participants and tasks. The bottom row represents the interaction between full outsourcing and deception (i.e., the difference between the differences in the rows above). Main numbers are posterior medians, numbers in the square brackets are 95% credible intervals.

@@

The effects of outsourcing to AI varied across the different tasks, especially 
for perceptions of warmth and morality (@fig-treatments-tasks-study1). For 
example, people who used AI for social tasks, such as writing an apology letter
or a bereavement card, were perceived as less warm, less moral, and lazier 
compared to people who completed the task themselves. This was true regardless
of whether AI was used as a tool or fully delegated to and whether the person
was honest or deceptive about their use of AI. By contrast, we observed weaker
effects of outsourcing for non-social tasks like writing computer code or 
planning a syllabus.

@@

In this study, we manipulated whether people used AI as a collaborative tool or 
copied the AI's output verbatim. In line with our predictions, we found that 
"fully" outsourcing to AI was perceived more negatively than using AI as a 
collaborative tool, particularly for socio-relational tasks. These negative 
perceptions of people who fully outsource their work to AI may be driven in part
by a general suspicion in unsupervised AI output due to what has been termed 
"algorithmic aversion" -- the tendency for people to distrust AI relative to 
humans [@Dietvorst2015]. This aversion is particularly strong in the moral 
domain [@Bigman2018]. Our results suggest that people may not only harbour 
negative perceptions of AIs themselves, but also of humans who unthinkingly copy
the output from these AIs, especially for socio-relational tasks.

We also manipulated whether people were honest or deceptive in their use of AI.
We found that people were seen as less moral and less trustworthy if they did
not acknowledge their use of AI when asked how they came up with their ideas.
This builds on previous work showing that people dislike and distrust those who
lie [@Tyler2006]. However, passing off AI-generated work as one's own is a 
specific kind of lie -- it involves taking undeserved credit for the work of a 
machine and misrepresenting one's own emotional capabilities and writing skills.
Interestingly, we found that failing to acknowledge use of AI had negative 
effects on perceptions of morality and trustworthiness across all the tasks we 
studied, suggesting that this phenomenon generalises across a range of 
socio-relational and professional tasks.

We have shown outsourcing to AI leads to negative character evaluations, 
especially when the AI output is copied verbatim and the work is passed off as 
one's own. However, it remains unclear whether these negative perceptions also
extend to the work itself. It could be, for example, that someone is perceived
badly for using ChatGPT to write their bereavement card, but the actual quality
of the writing is perceived to be just as good, if not better, than if the
person had written the card themselves. Indeed, evidence suggests that text
generated by ChatGPT is rated as higher quality than human-written text
[@Noy2023]. However, because this previous study used treatment-blind raters, it
was unable to assess whether knowing that a piece of text was AI-generated
affects how people perceive the text, while holding the quality of the text
constant. We explored this possibility in Study 2.

@@

We randomly allocated participants to one of six conditions in a 3x2 
between-subjects design. We manipulated the type of outsourcing: (*i*) no 
outsourcing control, (*ii*) using AI as a tool, and (*iii*) fully outsourcing to
AI. We also manipulated whether the task prompt was social or non-social.

@@

We told participants that they would read and evaluate a short piece of writing
from "another participant" (in reality, we generated the writing using ChatGPT 
version 4o and edited the text to appear more human-like). The prompt for the 
piece of writing varied between conditions:

@@

We then presented participants with a randomly-chosen pre-generated essay answer
to the prompt (see Supplementary Tables [-@supptbl-essay-answers-social-study2] 
and [-@supptbl-essay-answers-nonsocial-study2] for full essay answers). In the 
social conditions, the answer either referred to the participant's father, their
sister, or their best friend. In the non-social conditions, the answer either
referred to the book The Hobbit, the TV show Buffy the Vampire Slayer, or the
film Titanic. Reading times and responses to a follow-up comprehension question
suggested that participants read the essay answers in sufficient detail (see 
@supptbl-essay-comprehension-study2).

We asked participants about their perceptions of the essay answer and the "other
participant". We asked how well-written, meaningful, and authentic they thought
the answer was (7-point Likert scales), what letter grade they would give the
answer (A-E), and how much they would hypothetically reward the other
participant for their work (from £0.00 to £1.00). We also asked how well each of
the following words described the other participant: competent, warm, moral,
lazy, and trustworthy (7-point Likert scales).

At the end of the study, we gave participants a manipulation check and asked
them whether they believed the manipulation. Responses to these questions
suggested that the outsourcing manipulation was successful (see 
@supptbl-manipulation-check-study2). We also asked participants several 
questions about ChatGPT (see @suppfig-chatgpt-study2).

@@

We fitted two Bayesian multilevel models to the data. The first model was a 
multivariate cumulative-link ordinal model including all Likert scales as 
separate response variables. The second model was a zero-one-inflated-beta model
applied specifically to the reward question, which was a slider scale varying
between 0 and 1. For both models, we included fixed effects for the interaction
between outsourcing type and task type and varying intercepts and slopes for
essay answers. We used regularising priors for all parameters to impose 
conservatism on parameter estimates (see Supplementary Materials for full model 
specifications). All models converged normally ($\hat{R}$ ≤ 1.01).

@@

We found that essay responses that were ostensibly generated using AI were 
perceived as less meaningful and less authentic, but just as well-written, 
compared to essay responses ostensibly written by a human
(@fig-treatments-work-study2; @tbl-treatment-diffs-work-study2). Participants 
also marked AI-generated essays with a lower grade and rewarded AI-generated 
essays with a lower hypothetical monetary bonus.

@@

We did not find differences in perceptions of the work between the tool 
outsourcing and full outsourcing conditions, except for the reward question,
where fully outsourced essays (i.e., essays copied verbatim from ChatGPT) were
rewarded £0.23 less than essays generated using AI as a collaborative tool. We
did not find any differences between social and non-social tasks and did not
find any interaction effects.

Turning to character evaluations, we found that people were perceived more
negatively across all character evaluations if they outsourced the writing task
to AI, either by using ChatGPT as a collaborative tool or by copying the text
from ChatGPT verbatim (@suppfig-treatments-person-study2; 
@supptbl-treatment-diffs-person-study2). However, we did not find any 
differences in character evaluations between the tool outsourcing and full 
outsourcing conditions. We did not find differences in character evaluations 
between social and non-social tasks and did not find any interaction effects.

@@

Extending our previous results, we found that not only are people perceived 
negatively for outsourcing their work to AI, but the work produced by AI is 
itself also perceived negatively. In particular, text that has purportedly been
generated using AI is perceived to be less meaningful, less authentic, worthy of
a lower grade, and worthy of a lower monetary reward compared to human-generated
text. This is despite the actual text itself being held constant, perhaps
explaining why perceived writing quality did not consistently differ between
conditions.

Unlike in Study 1, we found no differences in the effect of AI-outsourcing
between social and non-social tasks. This may be due to the particular tasks we
chose. Writing *about* someone close to you is not quite the same as writing 
something *for* someone close to you, as is the case with wedding vows, love 
letters, and bereavement cards. We also found no differences between the tool
and full outsourcing conditions, aside from the lower rewards given to
participants in the latter condition. It is possible that differences between
conditions were smaller under this design and we were not sufficiently powered
to detect them.

>>>>

---
title: "Negative Perceptions of Outsourcing to Artificial Intelligence"
shorttitle: "Outsourcing to AI"
author:
  - name: Scott Claessens
    corresponding: false
    orcid: 0000-0002-3562-6981
    email: scott.claessens@gmail.com
    affiliations:
      - name: University of Kent
        department: School of Psychology
        address: Keynes College
        city: Canterbury
        country: UK
        postal-code: CT2 7NP
    role:
      - conceptualization
      - data curation
      - formal analysis
      - investigation
      - methodology
      - visualization
      - writing
      - editing
  - name: Pierce Veitch
    corresponding: false
    orcid: 0009-0005-3364-7470
    email: pv201@kent.ac.uk
    affiliations:
      - name: University of Kent
        department: School of Psychology
        address: Keynes College
        city: Canterbury
        country: UK
        postal-code: CT2 7NP
    role:
      - formal analysis
      - methodology
      - editing
  - name: Jim A.C. Everett
    corresponding: true
    orcid: 0000-0003-2801-5426
    email: j.a.c.everett@kent.ac.uk
    affiliations:
      - name: University of Kent
        department: School of Psychology
        address: Keynes College
        city: Canterbury
        country: UK
        postal-code: CT2 7NP
    role:
      - conceptualization
      - funding acquisition
      - methodology
      - supervision
      - editing
author-note:
  status-changes: 
    affiliation-change: ~
    deceased: ~
  disclosures:
    study-registration: ~
    data-sharing: All data and original code can be found on GitHub at <https://github.com/ScottClaessens/outsourcing>.
    related-report: ~
    conflict-of-interest: The authors have no conflicts of interest to disclose.
    financial-support: This work was generously supported by funding from the Horizon Europe UK Guarantee via the UKRI (EP/Y00440X/1) and a Philip Leverhulme Prize (PLP-2021-095) awarded to JACE.
    gratitude: ~
    authorship-agreements: ~
abstract: >
  As artificial intelligence (AI) tools become increasingly integrated into 
  daily life, people are beginning to outsource not only professional tasks but 
  also socio-relational ones. Large language models like ChatGPT can generate 
  wedding vows, speeches, and personal messages, raising questions about how 
  individuals who use AI for such tasks are perceived by others. In this paper,
  we conduct five pre-registered studies with British participants (N = 3,649)
  to understand how people view those who outsource tasks to AI, and how this
  depends on how socio-relational the task is, whether AI is used as a tool or
  fully delegated to, and the acknowledgment of the AI use. We find negative
  perceptions of outsourcing, particularly for socio-relational tasks. We show
  that outsourcing makes us think more negatively about not only the person and
  their motivations, but also the outsourced work itself. Moreover, we provide
  insight into why this occurs: the reduced effort from outsourcing
  socio-relational tasks to AI signals that the output is less authentically 
  one's own and that the person cares less about the task. Our research 
  highlights the way that AI use shapes our perceptions of people, raising key
  philosophical questions about efficiency, authenticity, and social ties in a
  world filled with AI-mediated interactions.
keywords: [artificial intelligence, person perception, outsourcing, trust]
bibliography: bibliography.bib
engine: knitr
format:
  apaquarto-pdf: default
execute:
  echo: false
  warning: false
  error: false
crossref:
  custom:
    - kind: float
      key: suppfig
      latex-env: suppfig
      reference-prefix: Supplementary Figure
      space-before-numbering: true
      latex-list-of-description: Supplementary Figure
      caption-location: bottom
    - kind: float
      key: supptbl
      latex-env: supptbl
      reference-prefix: Supplementary Table
      space-before-numbering: true
      latex-list-of-description: Supplementary Table
      caption-location: top
floatsintext: true
---

@@

The widespread release of generative AI language models has transformed daily 
life, offering the potential to perform a variety of tasks more efficiently and,
in some cases, with greater effectiveness than by doing them oneself. But as AI
becomes more widely available, people are not only using it to assist them with
things like preparing dinner recipes, writing data analysis code, and planning
daily schedules. Increasingly, AI might be used beyond routine or technical
domains to instead assist in tasks that are more socio-relational in nature,
like writing wedding vows, apology notes, and love letters. Anecdotal evidence
suggests that not only is AI-outsourcing of this kind already happening, but
that it potentially has serious effects on how we judge others. In a recent
Reddit post, a disgruntled newlywed tells the story of her husband using ChatGPT
to write his wedding vows, expressing her discomfort with outsourcing something
to AI that, to her, is deeply meaningful and a reflection of their love for one 
another [@miramar0]. Outsourcing tasks -- especially socio-relational ones -- to
AI tools may be efficient, but could have negative consequences for person 
perception.

There is nothing new, in principle, about outsourcing tasks. For hundreds of 
years, personal assistants have organised daily schedules, recipe-books have 
provided meal plans, and guidebooks have created travel itineraries. In the 
socio-relational domain, ghostwriters have long-existed, and the internet is 
abound with professional paid services for writing wedding vows and personal 
speeches. AI merely supercharges what is an ancient human impulse: the push to 
reduce mental energy by outsourcing parts of our work onto people, books, tools,
or systems. But even if outsourcing is an old phenomenon, the rapid shift in 
availability and use of AI models has fundamentally changed the ease with which
people can outsource work, what kinds of tasks they can outsource, and the way
in which they can outsource. These new developments in society mean that even as
an old phenomenon in new clothes, there is much we still need to know about
outsourcing.

First, we need to know how people who outsource tasks to AI are perceived. We 
know that people are increasingly using large language models (LLMs) for a wide
variety of tasks [@UKGov2024]. Due to their ubiquity, perhaps outsourcing to
LLMs might not lead to negative perceptions? We are sceptical. We know that
people dislike it when others "free ride" or reduce effort while benefiting from
collective resources [e.g. @Cubitt2011; @Kerr1983] and that people's outputs are
perceived as more valuable the more effort was ostensibly put into them
[@Kruger2004]. Moreover, exertion of effort is deemed morally admirable and is 
rewarded, even in situations where effort does not directly generate additional
product, quality, or economic value, suggesting that effort itself is moralised
[@Celniker2023]. Given this, even if AI tools are widely available and pitched
as improving efficiency, the core social psychological processes are likely to 
remain: someone is expending less effort to achieve a task, and people value 
effort. Indeed, some work shows that describing someone as using AI for a 
relational task led to the perception they expended less effort and were less 
satisfied with their relationship [@Liu2024] and other unpublished work looking
at perceptions of people using AI to complete academic assignments finds that 
using AI leads to more negative perceptions of moral character and suitability 
as a partner [@Roth2025].

Second, we need to know whether the *type* of task that people are outsourcing 
matters. One might expect outsourcing to be perceived negatively regardless of 
the type of task being outsourced -- if effort is generally moralised, then the 
domain in which it is expended (or not) should have little impact. However, 
there are reasons to expect differences between social tasks like writing vows 
and non-social tasks like writing computer code. We know that different norms,
standards, and expectations can be applied to social and non-social tasks and 
exchanges [e.g. @Fiske1992; @Heider1958; @Malle2022]. Moreover, from a 
philosophical perspective, it often matters not only *whether* something is 
done, but *how* it is done [@Stohr2006]. An apology is not just about hearing 
someone say "I am sorry", but seeing genuine regret; a love letter is not just 
about hearing someone say "I love you", but seeing depth of emotion; and a 
bereavement letter is not just about hearing someone say "I am sorry for your 
loss", but seeing an understanding for the powerful human experience of loss.
There is, perhaps especially for social tasks, value not only in the outcome of 
doing something, but the *process* too [@Goodman2010]. To understand any 
potential negative effects of outsourcing to AI, we must therefore look at a 
broad range of non-social and social tasks, rather than draw broad conclusions
based on a few use cases.

Third, we need to know how different ways of outsourcing to AI influence 
negative perceptions. Someone who "fully" outsources a task to AI by simply 
giving it a prompt and copying the output word-for-word might be perceived very
differently to someone who gives the AI a prompt, revises the work accordingly,
and finishes it themselves -- using AI as a *collaborative tool*, rather than as
a replacement. Similarly, someone could deceive others about their use of AI or
be perfectly honest about it. While it seems reasonable to assume that "fully"
outsourcing would be perceived worse than using AI as a tool, and that not
acknowledging AI use would be perceived worse than being honest about it, it
remains unclear how much this reduces negative perceptions: if someone uses AI
in the "best" way, by using it as a collaborative tool and being open about this
use, would they still suffer negative social consequences from doing so?

Fourth, we need to know how outsourcing to AI, in different kinds of tasks and 
in different ways, may shape different *kinds* of social perceptions. People can
judge others on separate dimensions of warmth and competence [e.g. @Abele2021;
@Fiske2007] as well as on dimensions of morality and trustworthiness
[@Goodwin2014]. It remains unclear how outsourcing to AI might lead to 
differential character judgments across these different dimensions.

Fifth, we need to understand *why* outsourcing to AI, and therefore expending 
less effort, might have these effects. Previous work has focused on how 
expending less effort leads to negative perceptions of others [@Celniker2023].
But this raises the question of *why* effort is seen as important and what
exactly it is signalling to others, beyond one's general cooperative intent. It
is possible that outsourcing leads to negative perceptions because the lack of
effort spent on the task signals something more fundamental about how authentic
one is and how much one cares about the task: when someone chooses to outsource
a love letter to an AI, they might be seen as valuing that love letter and what
it represents less. It could be this second-step order of perceptions that is
the key driver of negative perceptions, especially for socio-relational tasks.

## Present Research

In this paper, we build on classic social psychological work on character 
inferences from reduced effort to understand how people view others who
outsource different kinds of tasks, in different ways, for different reasons, to
AI. Across five pre-registered experiments with British participants, we seek
not only to understand how reduced effort through AI-outsourcing might shape
perceptions of others, but also to understand in more depth *why* it is that
reduced effort has the effect that it does.

In our initial pilot studies to motivate this work, we found that people who
outsource a range of tasks to AI or another person are perceived more negatively
than people who complete the tasks by themselves (see Supplementary Materials).
In Study 1, we look at the effects of task type, AI use, and honesty. We explore
how people perceive others who outsource different kinds of tasks with different
levels of social relevance (e.g., from daily schedules, computer code and dinner
recipes to wedding vows, apology letters, or bereavement cards), manipulating
whether people use AI as a collaborative tool or "fully" outsource to AI and 
whether they are honest or deceptive about their use of AI. After turning to
look at perceptions of both outsourcers and the outsourced work in Study 2, in
Studies 3-5 we probe why outsourcing may have negative effects on how we
evaluate others. In Study 3, we test potential mechanisms of perceived effort
and authenticity by looking at how people evaluate others who either spend a lot
or little time crafting the AI prompts, and who either outsource to a generic or
personalised AI. In Study 4, we test the potential mechanism of perceived
importance in the task by manipulating people's reasons for using AI -- either 
because they wanted to save time or because they cared about the task and
thought that AI would improve their work. Finally, in Study 5, we bring these 
different potential mechanisms together to explore the different pathways that 
influence the relationship between outsourcing and negative perceptions, 
focusing on perceived effort, authenticity, and care in the task.

@@

Ethical approval was granted for all studies in this paper by the University of
Kent's Psychology Research Ethics Panel. Participants in all studies provided
informed consent and were debriefed after the study.

@@

We fitted Bayesian multivariate multilevel cumulative-link ordinal models to the
data using the *brms* R package [@Burkner2017]. We modelled each character 
evaluation -- competence, warmth, morality, laziness, and trustworthiness -- as 
a separate response variable and included fixed effects for conditions, varying 
intercepts for participants, and varying intercepts and slopes for tasks. We 
used regularising priors for all parameters to impose conservatism on parameter 
estimates. All models converged normally ($\hat{R}$ ≤ 1.01).

@@

We first looked at the overall results averaging over tasks. Across all five 
character evaluations, we found that fully outsourcing to AI (i.e., copying the 
AI output verbatim) was perceived more negatively than using AI as a 
collaborative tool (@fig-treatments-study1; @tbl-treatment-diffs-study1). By 
contrast, we found that deception about AI usage had specific negative effects 
on perceptions of morality and trustworthiness: people who did not acknowledge 
their use of AI were perceived as less moral and less trustworthy. We did not 
find any interaction effects between full outsourcing and deception.

```{r, fig.height=6, fig.width=6}
#| label: fig-treatments-study1
#| fig-cap: Overall Character Evaluations in Study 1
#| fig-align: center
#| apa-note: Participants in the control condition and four AI outsourcing conditions evaluated people in the scenarios on (a) competence, (b) warmth, (c) morality, (d) laziness, and (e) trustworthiness. Coloured points represent participant responses to the questions, jittered for easier viewing. Black points are estimated marginal means from the fitted model, pooling over participants and tasks. Black points and line ranges represent posterior medians and 95% credible intervals, respectively.

@@

```{r}
#| label: tbl-treatment-diffs-study1
#| tbl-cap: Overall Pairwise Contrasts in Study 1 \vspace{20pt}
#| apa-note: Numbers reflect differences in marginal means on a 7-point Likert scale, pooling over participants and tasks. The bottom row represents the interaction between full outsourcing and deception (i.e., the difference between the differences in the rows above). Main numbers are posterior medians, numbers in the square brackets are 95% credible intervals.

@@

The effects of outsourcing to AI varied across the different tasks, especially 
for perceptions of warmth and morality (@fig-treatments-tasks-study1). For 
example, people who used AI for social tasks, such as writing an apology letter 
or a bereavement card, were perceived as less warm, less moral, and lazier 
compared to people who completed the task themselves. This was true even if the 
person used AI as a tool and was honest about their usage of AI. By contrast, we 
observed weaker effects of outsourcing for non-social tasks like writing 
computer code or planning a syllabus.

@@

In Study 1, we looked at how people who outsourced to AI in different ways were 
perceived across a broad range of social and non-social tasks. In line with our 
predictions, we found that "fully" outsourcing to AI was perceived more 
negatively than using AI as a collaborative tool, particularly for 
socio-relational tasks. We also found, predictably, that people were seen as 
less moral and less trustworthy if they did not acknowledge their use of AI. 
Importantly, though, we show that even using AI in the "best" way -- only as a 
tool and being honest about one's usage -- still led to negative social 
perceptions for the more socio-relational tasks like writing a love letter, an 
apology, or wedding vows.

In Study 2, we investigate whether these negative perceptions extend to the work 
itself and remain after seeing the output. It could be, for example, that 
someone is perceived badly for using ChatGPT to write their bereavement card, 
but the writing itself is seen as equally well-written and authentic, if not 
more so, than if the person had written the card themselves. Indeed, evidence 
suggests that text generated by ChatGPT is rated as higher quality than 
human-written text [@Noy2023]. Moreover, it is possible that seeing appropriate 
output could mitigate negative perceptions by highlighting how the AI can in 
fact perform the task well. We explored these possibilities in Study 2.

@@

We randomly allocated participants to one of six conditions in a 3x2 
between-subjects design. We manipulated the type of outsourcing: (*i*) no 
outsourcing control, (*ii*) using AI as a tool, and (*iii*) fully outsourcing to
AI. Here, in contrast to Study 1, we also explicitly manipulated whether the 
task prompt was social or non-social.

@@

We told participants that they would read and evaluate a short piece of writing 
from "another participant". In reality, we generated the writing using ChatGPT 
version 4o. We asked ChatGPT to generate a 300 word response to the prompt and 
to write convincingly like a real human. We then edited the text to appear more 
human-like by, for example, removing classic AI markers like dashes and 
concluding sentences and ensuring that the information was not too generic, such 
that the writing could reasonably be attributed to both a human and AI.

The prompt for the piece of writing varied between conditions:

@@

Next, we presented participants with a randomly-chosen pre-generated essay 
answer to the prompt (see Supplementary Tables 
[-@supptbl-essay-answers-social-study2] and 
[-@supptbl-essay-answers-nonsocial-study2] for full essay answers). In the 
social conditions, the answer either referred to the participant's father, their
sister, or their best friend. In the non-social conditions, the answer either
referred to the book The Hobbit, the TV show Buffy the Vampire Slayer, or the
film Titanic. Reading times and responses to a follow-up comprehension question
suggested that participants read the essay answers in sufficient detail (see 
@supptbl-essay-comprehension-study2).

Finally, we asked participants about their perceptions of the essay answer and 
the "other participant". We asked how well-written, meaningful, and authentic 
they thought the answer was (7-point Likert scales), what letter grade they 
would give the answer (A-E), and how much they would hypothetically reward the 
other participant for their work (from £0.00 to £1.00). We also asked how well 
each of the following words described the other participant: competent, warm, 
moral, lazy, and trustworthy (7-point Likert scales).

At the end of the study, we gave participants a manipulation check and asked
them whether they believed the manipulation. Almost all participants correctly 
reported the condition that they were in and most participants stated that they 
believed the essay response was written in the way we described, suggesting that 
the manipulation was successful (see @supptbl-manipulation-check-study2). We 
also asked participants several questions about ChatGPT.

@@

We fitted two Bayesian multilevel models to the data. The first model was a 
multivariate cumulative-link ordinal model including all Likert scales as 
separate response variables. The second model was a zero-one-inflated-beta model
applied specifically to the reward question, which was a slider scale varying
between 0 and 1. For both models, we included fixed effects for the interaction
between outsourcing type and task type and varying intercepts and slopes for
essay answers. We used regularising priors for all parameters to impose 
conservatism on parameter estimates. All models converged normally 
($\hat{R}$ ≤ 1.01).

@@

We first looked at character evaluations. We found that even when provided with 
concrete output, people were still perceived more negatively across all 
character evaluations if they outsourced the writing task to AI, either by using 
ChatGPT as a collaborative tool or by copying the text from ChatGPT verbatim 
(@suppfig-treatments-person-study2; @supptbl-treatment-diffs-person-study2). 
In contrast to Study 1, however, we did not find any differences in character 
evaluations between the tool outsourcing and full outsourcing conditions. We did 
not find differences in character evaluations between social and non-social 
tasks and did not find any interaction effects.

Turning to evaluations of the work itself, we found that the AI-outsourced work 
(either outsourced by using AI as a collaborative tool or fully outsourced) was 
judged as being equally well written to the work in the control condition 
(@fig-treatments-work-study2; @tbl-treatment-diffs-work-study2). This is in line 
with the writing indeed being identical in all conditions. Interestingly, 
however, we found that essay responses that were ostensibly generated using AI 
were perceived as less meaningful and less authentic compared to essay responses 
ostensibly written by a human. Participants also marked AI-generated essays with 
a lower grade and rewarded AI-generated essays with a lower hypothetical 
monetary bonus. In contrast to Study 1, we did not find differences in 
perceptions of the work between the tool outsourcing and full outsourcing 
conditions, except for the reward question, where fully outsourced essays (i.e., 
essays copied verbatim from ChatGPT) were rewarded £0.23 less than essays 
generated using AI as a collaborative tool. We did not find any differences 
between social and non-social tasks and did not find any interaction effects.

@@

@@

In Study 2, we turned to look at how people perceived both the outsourcer and 
the outsourced work when given specific output in a social or non-social task 
that was described as being produced independently by a person, produced by a 
person in collaboration with AI as a tool, or outsourced in full to AI. We find 
that our results generalise from character judgments to perceptions of the work 
itself: text purportedly generated using AI was perceived to be less meaningful, 
less authentic, and less reward-worthy compared to the same text described as 
human-generated.

Surprisingly, we found no differences in the effect of AI-outsourcing between 
social and non-social tasks. This may be due to the particular tasks we chose. 
Writing *about* someone close to you is not quite the same as writing something 
*for* someone close to you, as is the case with wedding vows, love letters, and 
bereavement cards. We also found no differences between the tool and full 
outsourcing conditions, aside from the lower rewards given to participants in 
the latter condition. It is possible that because the set-up described to 
participants was of another participant who was asked to produce work on 
Prolific and then admitted they used AI, participants saw any kind of AI use as 
violating an implicit contract between the survey requester and respondent and 
judged them negatively accordingly.
